---
title: "Mapping in R"
date: "`r paste0(format(Sys.time(), '%d %B %Y'))`"
author: Kevin Cazelles, post-doc, McCann Lab
# lang: french
output:
    ioslides_presentation:
        incremental: true
        widescreen: true
        #logo: ./img/logo.png
---





# Why?

## Why use R as a GIS?

### 1- Workflow

### 2- Quite efficient

### 3- Interface

<!-- 5 years it has evolved at a very fast pace... -->



## Workflow

### **R for everything** (*almost*)

- import your data
- format your data
- analyze your data
- visualize your data
- export your data
- create your own function/package



## Efficiency

- well-defined classes
- many formats (+ convert different file)
- topology operations     
- statistical analyses
- visualization   


## Interface

- C/C++: [Geos](http://trac.osgeo.org/geos), [proj4](http://proj4.org/), [Gdal](http://www.gdal.org/)
- JavaScript: [Leaflet](https://rstudio.github.io/leaflet/)
- otherGIS: [RQGIS](https://cran.r-project.org/web/packages/RQGIS/index.html), [rgrass7](https://cran.r-project.org/web/packages/rgrass7/index.html), [RgoogleMaps](https://cran.r-project.org/web/packages/RgoogleMaps/index.html)
- access data: [osmdata](https://github.com/ropensci/osmdata)





# Overview

## Turning R into a powerful GIS

### 1- Define classes

- `sp`
- `raster`


### 2- Topology operations

- `rgeos`

### 3- Import / export

- `rgdal`


## Turning R into a powerful GIS

### 5- Analyses

- `spdep`
- `dismo`

### 4- Creating and editing maps

- `graphics`  
- `ggplot`
- `interactive plot`  

### 6- But

- Geo-referencing ...




# Package `sp`

## Classes

Classes / Functions         | Contents
--------------------------- | ---------------
Points                      | list of points (set of coordinates)
SpatialPoints               | list of points + CRS
SpatialPointsDataPoints     | list of points + CRS + attribute table
Line                        | a line (set of coordinates)
Lines                       | list of lines
SpatialLines                | list of lines + CRS
SpatialLinesDataFrame       | list of lines + CRS + attribute table


## Example: `SpatialPointsDataPoints`

```{r simpoint, echo = F, results='asis'}
library(knitr)
library(magrittr)
mydata <- data.frame(
  long = -82+2*runif(20),
  lat = 42+2*runif(20),
  var1 = rnorm(20),
  var2 = 10*runif(20)
)
kable(head(mydata))
```

## Example: `SpatialPointsDataFrame`

```{r sppoint}
library(sp)
mysp <- SpatialPointsDataFrame(
  coords = mydata[,1:2],
  data = mydata[,3:4],
  proj4string = CRS(
    "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0"
  )
)
```

## Example: `SpatialPointsDataFrame`

```{r sppoint2}
class(mysp)
isS4(mysp)
slotNames(mysp)
```

## Example: `SpatialPointsDataFrame`

```{r sppoint3}
mysp@proj4string
head(mysp@data)
```

## Change projection: `spTransform`

```{r spTransform}
(mysp2 <- spTransform(mysp, CRS=CRS("+proj=merc +ellps=GRS80")))
```


# Package `raster`


## Classes

### 1- SpatialGrid (package `sp`)

### 2- SpatialGrid (package `sp`)

### 3- RasterLayer (package `raster`)

### 4- RasterStack (package `raster`)

### 5- RasterBrick (package `raster`)


## Example:`raster`

```{r raster}
library(raster)
ras1 <- raster(matrix(runif(100*100,0,10),ncol=100,nrow=100),
    crs=CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"),
    xmn=-82, xmx=+80, ymn=+42, ymx=+44)
```

## Example:`raster`

```{r raster2}
ras1
```

## Retrieving free GIS data: `getData`

```{r getDat1}
head(getData("ISO3"))
```

## Example:`getData`

```{r getDat2}
## Country level:
mapBEL0 <- getData(name="GADM", country="BEL", path="./assets", level=0)
mapBEL1 <- getData(name="GADM", country="BEL", path="./assets", level=1)
tminW <- getData(name="worldclim", var="tmin", res=10, path="./assets")
mapBEL0
```

# Package `rgdal`

## Drivers

```{r rgdal1}
library(rgdal)
```

1. **writeOGR** / **writeGDAL**: to write spatial objects
2. **readOGR**/**readGDAL**: to read spatial files

## Drivers

```{r rgdal2}
head(ogrDrivers())
```

## Export

```{r rgdal3}
writeOGR(mysp, dsn="./assets", layer="mypoints",
    driver="ESRI Shapefile", overwrite_layer=TRUE)
```


## Import

```{r rgdal4}
mysp2 <- readOGR(dsn="assets/", layer="mypoints")
## Roads find on http://www.diva-gis.org/Data
canroads <- readOGR(dsn="assets/", layer="CAN_roads")
```

*Note*: there are also functions to import/export raster in `raster`





# Package `rgeos`

## Load the package

```{r rgeos}
library(rgeos)
```

## Load the package

```{r rgeos2}
buf <- gBuffer(mapBEL0, width=0.5)
diff <- gDifference(gBuffer(mapBEL0, width=0.5), gBuffer(mapBEL0, width=0.1))
```

# Package `mapview`

## Import package

```{r mapview}
library(mapview)
```

**NB:** it uses the leaflet package.

## Quick examples

```{r mapview2}
mapview(mysp, cex = 'var2')@map
```

## Quick examples

```{r mapview3}
mapview(mapBEL1)@map
```





## More

- https://insileco.github.io/Elections_fr-042017/

- https://insileco.github.io/fr-departement-election/




# Editing a map  


## A very basic map -- Shapefile

```{r basicmap}
plot(mapBEL0)
```

## A very basic map -- Raster

```{r basicmap2}
class(tminW)
```

## A very basic map -- Raster

```{r basicmap3}
plot(tminW)
```

## Cutomize a map -- Shapefile


```{r basicmap4, eval=F}
plot(mapBEL0, border='grey15', col='#E6E6E6', lwd=1.6)
plot(mapBEL1, lty=2, lwd=0.9, add=T)
points(4.3513, 50.8471, pch=19, col="#27df9d")
text(4.3513, 50.8471, text="Brussel", pos=3)
```

## Cutomize a map -- Shapefile


```{r basicmap5, echo=F}
plot(mapBEL0, border='grey15', col='#E6E6E6', lwd=1.6)
plot(mapBEL1, lty=2, lwd=0.9, add=T)
points(4.3513, 50.8471, pch=19, col="#27df9d")
text(4.3513, 50.8471, label="Brussel", pos=3)
```


## Cutomize a map -- Shapefile of roads

```{r mapview3b, echo=F}
# mapview(canroads)@map
plot(c(-82,-80), c(42,44), type='n', asp=1.5)
plot(canroads, add=T)
```






# Resources

## Useful links

- task view: https://cran.r-project.org/web/views/Spatial.html
- blog R-spatial: http://r-spatial.org/
- a nice introduction: http://rspatial.org/intr/index.html
- [sf](http://r-spatial.github.io/sf/)
- [create shapefile on line](http://geojson.io/)
- [free data at country level](http://www.diva-gis.org/gdata)
- find more data with [this list](https://freegisdata.rtwilson.com/)




<!-- ## cool

<div class="centered">
This text is centered.
</div>

## cool2

<div class="columns-2">

  - Bullet 1
  - Bullet 2
  - Bullet 3
</div> -->
