---
title: "Mapping in R"
date: "`r paste0(format(Sys.time(), '%d %B %Y'))`"
author: 'Kevin Cazelles, post-doc, McCann Lab, [<i class="fa fa-github" aria-hidden="true"></i>](https://github.com/KevCaz/mapsWithR)'
# lang: french
output:
  ioslides_presentation:
    incremental: true
    widescreen: true
    includes:
      in_header: kcheader.html
      #logo: ./img/logo.png
---


# In a nutshell

## Packages

- [`sf`](https://cran.r-project.org/web/packages/sf/index.html) vector files
- [`raster`](https://cran.r-project.org/web/packages/raster/index.html) raster
- [`mapview`](https://cran.r-project.org/web/packages/mapview/index.html) visualize

## Resources

- https://insileco.github.io/tuto/rinspace/rinspace_homepage/
- https://geocompr.robinlovelace.net/
- http://rspatial.org/intr/index.html
- https://www.r-spatial.org/
- http://geojson.io/#map=2/20.0/0.0
- https://freegisdata.rtwilson.com/



# Why?


## Mapping in Ecology?

1. show where your plots are

2. show how variables are distributed spatially

3. spatial analyses

4. many, many tools [<i class="fa fa-external-link" aria-hidden="true"></i>](http://www.nature.com/news/data-visualization-science-on-the-map-1.17024)


## R as a GIS?

1. Workflow

2. Quite efficient

3. Interface

<!-- 5 years it has evolved at a very fast pace... -->



## Workflow

### **R for everything** (*almost*)

1. import your data
2. format your data
3. analyze your data
4. visualize your data
5. export your data
6. create your own functions/packages



## Efficiency

1. well-defined classes
2. many formats (+ convert different file)
3. operations on geometries
4. statistical analyses
5. visualization


## Interface

1. C/C++: [Geos](http://trac.osgeo.org/geos), [proj4](http://proj4.org/), [Gdal](http://www.gdal.org/)
2. JavaScript: [Leaflet](https://rstudio.github.io/leaflet/)
3. otherGIS: [RQGIS](https://cran.r-project.org/web/packages/RQGIS/index.html), [rgrass7](https://cran.r-project.org/web/packages/rgrass7/index.html), [RgoogleMaps](https://cran.r-project.org/web/packages/RgoogleMaps/index.html)
4. retrieve data: [osmdata](https://github.com/ropensci/osmdata)





# How? An overview

## Turning R into a powerful GIS

### 1- Define classes

- `sp`
- `raster`
- `sf`


### 2- Topology operations

- `rgeos`
- `sf`

### 3- Import / export

- `rgdal`
- `sf`


## Turning R into a powerful GIS

### 4- Analyses (few examples)

- `spdep`
- `spatial`
- `gstat`
- `dismo`

### 5- Creating and editing maps

- `graphics`
- `ggplot`
- `interactive plot / web-based plot`
- [a blog post about this](https://insileco.github.io/2015/12/04/list-of-r-packages-to-create-and-edit-plots/)


## Turning R into a powerful GIS

### 6- But...

- Geo-referencing  [<i class="fa fa-external-link" aria-hidden="true"></i>](http://robertmyles.github.io/2016/08/13/geo-reference-an-image-in-r/)
- Visualizing large spatial objects
- Watershed analysis




# Package `sp`

## Classes

Classes / Functions         | Contents
--------------------------- | ---------------
Points                      | list of points (set of coordinates)
SpatialPoints               | list of points + CRS
SpatialPointsDataPoints     | list of points + CRS + attribute table
Line                        | a line (set of coordinates)
Lines                       | list of lines
SpatialLines                | list of lines + CRS
SpatialLinesDataFrame       | list of lines + CRS + attribute table


## Example: `SpatialPointsDataPoints`

```{r simpoint, echo = F, results='asis'}
library(knitr)
library(magrittr)
mydata <- data.frame(
  long = -82+2*runif(20),
  lat = 42+2*runif(20),
  var1 = rnorm(20),
  var2 = 10*runif(20)
)
kable(head(mydata))
```

## Example: `SpatialPointsDataFrame`

```{r sppoint}
library(sp)
mysp <- SpatialPointsDataFrame(
  coords = mydata[,1:2],
  data = mydata[,3:4],
  proj4string = CRS(
    "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0"
  )
)
```

## Example: `SpatialPointsDataFrame`

```{r sppoint2}
class(mysp)
isS4(mysp)
slotNames(mysp)
```

## Example: `SpatialPointsDataFrame`

```{r sppoint3}
mysp@proj4string
head(mysp@data)
```

## Change projection: `spTransform`

```{r spTransform}
(mysp2 <- spTransform(mysp, CRS=CRS("+proj=merc +ellps=GRS80")))
```


# Package `raster`


## Classes

### 0- SpatialGrid / SpatialPixel (package `sp`)

### 3- RasterLayer (package `raster`)

### 4- RasterStack (package `raster`)

### 5- RasterBrick (package `raster`)


## Example: `raster`

```{r raster}
library(raster)
ras1 <- raster(matrix(runif(100*100,0,10),ncol=100,nrow=100),
    crs=CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"),
    xmn=-82, xmx=+80, ymn=+42, ymx=+44)
```

## Example: `raster`

```{r raster2}
ras1
```

## Retrieving free GIS data: `getData`

```{r getDat1}
head(getData("ISO3"))
```

## Example: `getData`

```{r getDat2}
## Country level:
mapBEL0 <- getData(name="GADM", country="BEL", path="./assets", level=0)
mapBEL1 <- getData(name="GADM", country="BEL", path="./assets", level=1)
tminW <- getData(name="worldclim", var="tmin", res=10, path="./assets")
mapBEL0
```

# Package `rgdal`

## Drivers

```{r rgdal1}
library(rgdal)
```

1. **writeOGR** / **writeGDAL**: to write spatial objects
2. **readOGR**/**readGDAL**: to read spatial files

## Drivers

```{r rgdal2}
head(ogrDrivers())
```

## Export

```{r rgdal3}
writeOGR(mysp, dsn="./assets", layer="mypoints",
    driver="ESRI Shapefile", overwrite_layer=TRUE)
```


## Import

```{r rgdal4}
mysp2 <- readOGR(dsn="assets/", layer="mypoints")
## Roads find on http://www.diva-gis.org/Data
canroads <- readOGR(dsn="assets/", layer="CAN_roads")
```

*NB*: there are function to import/export raster in `raster`





# Package `rgeos`

## Load the package

```{r rgeos}
library(rgeos)
```

## Load the package

```{r rgeos2}
buf <- gBuffer(mapBEL0, width=0.5)
diff <- gDifference(gBuffer(mapBEL0, width=0.5), gBuffer(mapBEL0, width=0.1))
```


# package `sf`

## Simple features?

> The package `sf`  [...] aims at succeeding `sp` in the long term.

> Simple features or simple feature access refers to a formal standard
(ISO 19125-1:2004) that describes how objects in the real world can be
represented in computers, with emphasis on the spatial geometry of these objects

https://cran.r-project.org/web/packages/sf/vignettes/sf1.html


## Importing spatial object

```{r sf1}
library(sf)
mysf <- st_read(dsn="assets/mypoints.shp")
```


## Importing spatial object

```{r sf2}
print(mysf)
```

## Importing spatial object

```{r sf3}
mysf2 <- st_as_sf(mysp)
```

## Simple manipulations

```{r sf4, echo = FALSE}
# http://spatialreference.org/
mysf2 <- st_transform(mysf, 3857)
```

## Simple manipulations

```{r sf5}
plot(mysf[1])
```

##

```r
plot(st_buffer(mysf[1], 0.1))
```

## Simple manipulations

```{r sf6, echo = FALSE}
plot(st_buffer(mysf[1], 0.1))
```

### `sf` or `sp`?

- https://insileco.github.io/2017/09/12/efficiency-of-spatial-intersects-in-r/

- https://github.com/r-spatial/sf




# Package `mapview`

## Import package

```{r mapview}
library(mapview)
```

**NB:** it uses the `leaflet` package which uses `sp`.


## 1 line of code examples

```{r mapview2}
mapview(mysp, cex = 'var2')@map
```

## Quick examples

```{r mapview3}
mapview(mapBEL1)@map
```

## More (using `leaflet`)

- https://insileco.github.io/Elections_fr-042017/

- https://insileco.github.io/fr-departement-election/




# Editing a map


## A very basic map -- Shapefile

```{r basicmap}
plot(mapBEL0)
```

## A very basic map -- Raster

```{r basicmap2}
class(tminW)
```

## A very basic map -- Raster

```{r basicmap3}
plot(tminW)
```

## Customize a map -- Shapefile

```{r basicmap4, eval=F}
plot(mapBEL0, border='grey15', col='#E6E6E6', lwd=1.6)
plot(mapBEL1, lty=2, lwd=0.9, add=T)
points(4.3513, 50.8471, pch=19, col="#27df9d")
text(4.3513, 50.8471, text="Brussel", pos=3)
```

## Customize a map -- Shapefile

```{r basicmap5, echo=F}
plot(mapBEL0, border='grey15', col='#E6E6E6', lwd=1.6)
plot(mapBEL1, lty=2, lwd=0.9, add=T)
points(4.3513, 50.8471, pch=19, col="#27df9d")
text(4.3513, 50.8471, label="Brussel", pos=3)
```


# Resources

## Useful links to use R as a GIS

- task view: https://cran.r-project.org/web/views/Spatial.html; also following
to this `https://cran.r-project.org/web/views/`

```R
install.packages("ctv")
install.views("Spatial")
```

- [Document par Nicolas Casajus](https://qcbs.ca/wiki/_media/gisonr.pdf)
- http://r-spatial.org/
- http://www.rspatial.org/
- [Tutorial on datacamp](https://www.datacamp.com/courses/spatial-analysis-in-r-with-sf-and-raster)

## Useful links to get data

- [free data at country level](http://www.diva-gis.org/gdata)
- [Quebec free data](http://mffp.gouv.qc.ca/le-ministere/acces-aux-donnees-gratuites/)
- [find more spatial data](https://freegisdata.rtwilson.com/)
- [create shapefile on line](http://geojson.io/)
- EPSG: [link1](http://spatialreference.org/); [link2](http://epsg.io/)













# Let's practice

## Installation

### 1- Have a look at https://github.com/r-spatial/sf

### 2- Install R packages

```R
install.packages(c("sf", "sp", "rgeos", "rgdal", "raster", "mapview"))
```

## Alexis's data

```{r import, echo = F}
knitr::read_chunk('assets/alexis.R')
```

```{r}
<<read_data>>
```

##

```{r}
<<make_sp>>
```

##

```{r}
<<plot_1>>
```

##

```{r}
<<plot_2>>
```

##

```{r}
<<get_data>>
```

##

```{r}
<<plot_3>>
```

##

```{r}
<<over>>
```

##

```{r}
<<region>>
```

##

```{r}
<<plot_3b>>
```

##

```{r}
<<ra_elv>>
```

##

```{r}
<<plot_4>>
```


##

```{r}
<<buffer>>
```

##

```{r}
<<plot_5>>
```

##

```{r}
<<new_data>>
```

##

```{r, eval=F}
<<plot_6>>
```

##

```{r, echo=F, fig.height = 6, fig.width =4}
<<plot_6>>
```

<!--
<div class="centered">
This text is centered.
</div>

## cool2

<div class="columns-2">

  - Bullet 1
  - Bullet 2
  - Bullet 3
</div> -->
