
#### Autre example





## Data frame with Ontario's lakes

- the `.csv` file below is available in the [<i class="fa fa-github" aria-hidden="true"></i>](https://github.com/KevCaz/mapsWithR) repository

```{r lake, message=F}
lakedf <- read.csv('assets/lakeOnt.csv')
head(lakedf)
```


## Turn it into `SpatialPointDataFrame`

```{r lake2, message=F}
library(sp)
lakesp <- SpatialPointsDataFrame(
  coords = lakedf[,3:4],
  data = lakedf[,c(1,4)],
  proj4string = CRS(
    "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0"
  )
)
```


## Your SpatialPointDataFrame


```{r lake3, message=F}
lakesp
```


## Retrieve altitude for Canada

```{r canalt, message=F}
altCAN <- getData(name="alt", country="CAN", path="./assets/")
altCAN
```

## Retrieve altitude for Canada

```{r canalt2, echo=F, message=F, warning=F}
mapview(altCAN)
```

## Retrieve provinces' boundaries

```{r mapbound}
bouCAN <- getData(country='CAN', level=1, path="./assets/")
bouCAN
```


## Retrieve provinces' boundaries

```{r mapbound2}
bouCAN@data[,1:8]
```


## Canada -- Ontario

```{r mapont, warning=F}
bouCAN[11,]
```

## Canada -- Ontario

```{r mapont2, echo=F, warning=F}
par(mar=c(0,0,0,0))
plot(bouCAN[11,])
```

## Canada -- Ontario

```{r mapont3, eval=F, warning=F}
plot(crop(altCAN, bouCAN[11,]@bbox))
par(mar=c(0,0,0,0))
```

## Canada -- Ontario

```{r mapont3b, echo=F, warning=F}
plot(crop(altCAN, bouCAN[11,]@bbox))
par(mar=c(0,0,0,0))
```


## Canada -- Ontario elevation

```{r elvont, eval=F, warning=F}
altONT <- rasterize(bouCAN[11,], crop(altCAN, bouCAN[11,]@bbox), mask=TRUE)
mapview(altONT)+mapview(lakesp)
```


## Canada -- Ontario elevation

```{r elvont2, echo=F, warning=F}
altONT <- rasterize(bouCAN[11,], crop(altCAN, bouCAN[11,]@bbox), mask=TRUE)
mapview(altONT)+mapview(lakesp)
```

## Canada -- Ontario elevation

```{r elvont3, eval=F, warning=F}
par(mar=c(0,0,0,0))
plot(altONT)
plot(lakesp, add=T)
```

## Canada -- Ontario elevation

```{r elvont4, echo=F, warning=F}
par(mar=c(0,0,0,0))
plot(altONT)
plot(lakesp, add=T)
```


<!--
par(mar=c(0,0,0,0))
plot(altONT)
plot(lakesp, add=T)
--------------------
south <- drawPoly()
south@proj4string <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0")
-----------------------
lakesp$North = 'N'
-----------------------
plot(altONT)
plot(lakesp[which(lakesp%over%south==1),], add=T)
------------------------
lakesp$North[which(lakesp%over%south==1)] = 'S'
------------------------
plot(altONT)
points(lakesp, pch=lakesp@data$North, cex=0.8)
------------------------
mKrig
packages fields
-->
